---
title: "Datasets"
author: "Fabian Strobel"
date: '2022-04-30'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

To get an overview the four given datasets are loaded into variables.

```{r}
# How I loaded the datasets (L.)

print(memory.size())

print(memory.limit())

memory.limit(size=56000) 


tcga_exp = readRDS(file.choose())

tcga_annot = readRDS(file.choose())

tcga_tumor_norm = readRDS(file.choose())

genesets = readRDS(file.choose())
```

```{r}
tcga_exp = readRDS("~/R/datasets/tcga_tumor_log2TPM.RDS")
tcga_annot = readRDS("~/R/datasets/tcga_tumor_annotation.RDS")
tcga_tumor_norm = readRDS("~/R/datasets/tcga_tumor_normal_datascience_proj_2022.RDS")
genesets = readRDS("~/R/datasets/hallmarks_genesets.rds")
```

The first 2 objects are data frames.
```{r}
dim(tcga_exp)
dim(tcga_annot)
```

The third object is a list with 5 entries corresponding to 5 tumor types (BRCA = breast cancer; KIRC = kidney renal carcinoma; LUAD = lung adenocarcinoma; PRAD = prostate adenocarcinoma; THCA = Thyroid carcinoma); each entry of the list is itself a list with 3 entries, corresponding to (1) a data frame with expression values for tumor samples, (2) a data frame with expression values for normal samples from the same patients, (3) a data frame with clinical annotations for these patients.
```{r}
prad = tcga_tumor_norm[["PRAD"]]
prad.tumor = prad[["tumor"]]
prad.norm = prad[["normal"]]
prad.annot = prad[["clinical"]]

dim(prad.tumor)
dim(prad.norm)
dim(prad.annot)
```

The last object is again a list
```{r}
class(genesets)
names(genesets)
names(genesets$genesets)

## these are the first genes of the first geneset
head(genesets$genesets[[1]])

## the element description contains a description of the genesets
head(genesets$description)
```