---
title: "01_pancancer_principle_component_analysis"
author: "Fabian Strobel"
date: '2022-06-06'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd('../')
directory <- "date/tcga_pancancer"
tcga_pancancer <- readRDS(directory)
```

Principal component analysis
```{r}
#pca <- prcomp(tcga_pancancer)
```

###
not transposed matrix
###
```{r}
pca <- readRDS("data/tcga_pancancer_pca")
```


Each PC explains part of overall variance in tcga matrix, proportional to corresponding eigenvalue

Analysis of variance
```{r}
variance = (pca$sdev)^2
barplot(log(variance), main = "log(variance) after pca for pan-cancer dataset", xlab = "PC_i", ylab = "log(variance)")
```

Scaling on 1 gives proportion of variance in percent
```{r}
prop.var = variance/sum(variance)

summary(prop.var)

sum(which(prop.var > quantile(prop.var, 0.999))) ## 55 values

head(prop.var, sum(which(prop.var > quantile(prop.var, 0.999))))

## Visualization
barplot(prop.var[1:sum(which(prop.var > quantile(prop.var, 0.999)))])
barplot(prop.var[2:sum(which(prop.var > quantile(prop.var, 0.999)))])

## The 55 first PCs explain 89.1 % of the total variance (PC1 -> 78.4 %!!!)
sum(prop.var[1:sum(which(prop.var > quantile(prop.var, 0.999)))])
```
Let's look whether we find clusters when plotting different PCs against each other:
```{r}
par(mfrow = c(2,3))

new.coordinates = pca$x

plot(new.coordinates[,1],new.coordinates[,2],pch=19,
     xlab='PC1',ylab='PC2')

plot(new.coordinates[,1],new.coordinates[,3],pch=19,
     xlab='PC1',ylab='PC3')

plot(new.coordinates[,1],new.coordinates[,4],pch=19,
     xlab='PC1',ylab='PC4')

plot(new.coordinates[,2],new.coordinates[,3],pch=19,
     xlab='PC2',ylab='PC3')

plot(new.coordinates[,2],new.coordinates[,4],pch=19,
     xlab='PC2',ylab='PC4')

plot(new.coordinates[,3],new.coordinates[,4],pch=19,
     xlab='PC3',ylab='PC4')

par(mfrow = c(1,1))
```


How to chose the "right" amount of PCs without risking to loose much information?
```{r}
## Kaiser rule (keep PCs with eigenvalue >= mean eigenvalue)

## Jottclife rule (keep PCs with eigenvalue >= 70% ofmean eigenvalue)

```


###
transposed matrix
###
```{r}
t_pca <- readRDS("data/tcga_pancancer_pca_transposed")
```


Each PC explains part of overall variance in tcga matrix, proportional to corresponding eigenvalue

Analysis of variance
```{r}
variance = (t_pca$sdev)^2
barplot(log(variance), main = "log(variance) after pca for pan-cancer dataset", xlab = "PC_i", ylab = "log(variance)")
```

Scaling on 1 gives proportion of variance in percent
```{r}
prop.var = variance/sum(variance)

summary(prop.var)

sum(which(prop.var > quantile(prop.var, 0.999))) ## 55 values

head(prop.var, sum(which(prop.var > quantile(prop.var, 0.999))))

## Visualization
barplot(prop.var[1:sum(which(prop.var > quantile(prop.var, 0.999)))])
barplot(prop.var[2:sum(which(prop.var > quantile(prop.var, 0.999)))])

## The 55 first PCs explain 52.2 % of the total variance
sum(prop.var[1:sum(which(prop.var > quantile(prop.var, 0.999)))])
```
Let's look whether we find clusters when plotting different PCs against each other:
```{r}
new.coordinates = t_pca$x

par(mfrow = c(2,3))

plot(new.coordinates[,1],new.coordinates[,2],pch=19,
     xlab='PC1',ylab='PC2')

plot(new.coordinates[,1],new.coordinates[,3],pch=19,
     xlab='PC1',ylab='PC3')

plot(new.coordinates[,1],new.coordinates[,4],pch=19,
     xlab='PC1',ylab='PC4')

plot(new.coordinates[,2],new.coordinates[,3],pch=19,
     xlab='PC2',ylab='PC3')

plot(new.coordinates[,2],new.coordinates[,4],pch=19,
     xlab='PC2',ylab='PC4')

plot(new.coordinates[,3],new.coordinates[,4],pch=19,
     xlab='PC3',ylab='PC4')

par(mfrow = c(1,1))
```