---
title: "Differentially expressed gene list"
date: '2022-06-18'
author: "Lottida Phondeth"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First load the data
```{r}
setwd('..')
tcga_tumor_norm <- readRDS("data/PRAD_TvsN.rds")

```

To visually grasp where the differentially expressed genes are coming from, the code for the volcano plot was applied a again 
create a data frame that contains the gene expression values for tumor and normal cell data
```{r}
PRAD_tumor_normal <- cbind(tcga_tumor_norm[["tumor"]], tcga_tumor_norm[["normal"]])

```
calculate the mean and the FC
```{r}
mean_t = apply(PRAD_tumor_normal[,1:52], 1, mean)
mean_n = apply(PRAD_tumor_normal[53:104], 1, mean)
e_mean_tumor <- 2^mean_t
e_mean_normal <- 2^mean_n
FoldChange <- e_mean_tumor / e_mean_normal
log2_FoldChange <- log2(FoldChange)
```

The wilcox test is used to determine the significance of our FC
```{r}
p_value <- apply(PRAD_tumor_normal, 1, function(x) wilcox.test(x[1:52], 
                                                         x[53:104], 
                                                         paired = TRUE,
                                                         exact = FALSE)$p.value)
log10_p_value <- log10(p_value)
```


Plot not corrected p-values
```{r}
hist(p_value, breaks = 20)
```


correct the p-values for multiple testing
```{r}
corrected_p_value <- p.adjust(p_value, method = "bonferroni")
hist(corrected_p_value, breaks = 20)
log10_corrected_p_value <- log10(corrected_p_value)
```
create a volcano plot with not corrected p-values
```{r}
dat <- data.frame(log2_FoldChange, log10_p_value)

# add a column of NAs
dat$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP" 
dat$diffexpressed[dat$log2_FoldChange > log2(2) & dat$log10_p_value < log10(0.05)] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
dat$diffexpressed[dat$log2_FoldChange < log2(0.5) & dat$log10_p_value < log10(0.05)] <- "DOWN"

# Change point appearance 
color <- c("blue", "red", "black")
names(color) <- c("DOWN", "UP", "NO")

library(ggplot2)

ggplot(data = dat, mapping = aes(x=log2_FoldChange, 
                                 y= -log10_p_value, 
                                 col = diffexpressed))+
  geom_point()+
  scale_color_manual(values = color)+
  geom_vline(xintercept=c(log2(0.5), log2(2)), linetype="dashed", col="black")+
  geom_hline(yintercept=-log10(0.05), linetype="dashed", col="black")+
  scale_x_continuous(breaks = c(seq(-7, 7, 2)),
                     limits = c(-7, 7))+
  labs(title = "Volcano plot")
```

plot volcano plot with corrected p-values
```{r}
dat_corrected <- data.frame(log2_FoldChange, log10_corrected_p_value)

# add a column of NAs
dat_corrected$diffexpressed <- "NO"
# if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP" 
dat_corrected$diffexpressed[dat_corrected$log2_FoldChange > log2(2) & dat_corrected$log10_corrected_p_value < log10(0.05)] <- "UP"
# if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
dat_corrected$diffexpressed[dat_corrected$log2_FoldChange < log2(0.5) & dat_corrected$log10_corrected_p_value < log10(0.05)] <- "DOWN"

# Change point appearance 
color <- c("blue", "red", "black")
names(color) <- c("DOWN", "UP", "NO")

library(ggplot2)

ggplot(data = dat_corrected, mapping = aes(x= log2_FoldChange, 
                                   y= -log10_corrected_p_value, 
                                   col= diffexpressed))+
  geom_point()+
  scale_color_manual(values = color)+
  geom_vline(xintercept=c(log2(0.5), log2(2)), linetype="dashed", col="darkgreen")+
  geom_hline(yintercept=-log10(0.05), linetype="dashed", col="darkgreen")+
  scale_x_continuous(breaks = c(seq(-7, 7, 2)),
                     limits = c(-7, 7))+
  labs(title = "Volcano plot")
```

create a gene list with differentially expressed genes

```{r}
n_df = which(dat_corrected$diffexpressed == "NO")
length(n_df) #6456 genes are not differentially expressed 
dat_df = dat_corrected[-n_df,] #we are left with 1345 genes
```

extract the names
```{r}
dat_df_names = as.list(rownames(dat_df))
```

reduce the PRAD_tumor_normal data frame
```{r}
PRAD_tumor_normal_df = subset(PRAD_tumor_normal,rownames(PRAD_tumor_normal) %in% dat_df_names)
```

Choose a heatmap color palette
```{r}
library(RColorBrewer)
heat_colors <- brewer.pal(n = 6, name = "YlOrRd")
```

generate a Heatmap
```{r}
library(pheatmap)
pheatmap(PRAD_tumor_normal_df,
         cluster_rows = T,
         show_rownames = F,
         show_colnames = F,
         fontsize = 8)
```

What is the correlation between the genes?
```{r}
PRAD_tumor_normal_df_cor = cor(t(PRAD_tumor_normal_df))
```

plot a heatmap of the correlation matrix of the genes
```{r}
library(pheatmap)
pheatmap(PRAD_tumor_normal_df_cor,
         cluster_rows = T,
         show_rownames = F,
         show_colnames = F,
         fontsize = 8)
```

