---
title: "fGSEA with Fold Change as ranks"
author: "Lottida Phondeth"
date: '2022-06-11'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First load the data

```{r}
setwd('..')
tcga_tumor_norm <- readRDS("data/PRAD_TvsN.rds")

# Fuse data frame PRAD normal with PRAD tumor
PRAD_tumor_normal <- cbind(tcga_tumor_norm[["tumor"]], tcga_tumor_norm[["normal"]])
```

First calculate the mean to be able to determine the fold change

```{r}
mean_n <- apply(PRAD_tumor_normal[ ,1:52], 1, mean)

mean_t <- apply(PRAD_tumor_normal[ ,53:104], 1, mean)
```

Calculate the Fold Change and the log2FoldChange

```{r}
e_mean_t <- 2^mean_t
e_mean_n <- 2^mean_n
FC <- e_mean_t / e_mean_n

# Scale Fold Change 
log2FC <- log2(FC)
```

The paired Wilcox-test is used to determine the significance of the FC

```{r}
p_value <- apply(PRAD_tumor_normal, 1, function(x) wilcox.test(x[1:52], 
                                                         x[53:104], 
                                                         paired = TRUE,
                                                         exact = FALSE)$p.value)

# Scale p-values 
neg_log10_p_value <- -log10(p_value)
```

Create a data frame with the scaled p value, FC and the gene ensID

```{r}
gene_ens_ID <- rownames(PRAD_tumor_normal)

dat_for_ranking <- data.frame(gene_ens_ID, neg_log10_p_value, log2FC, FC)
```

Let's check whether there are any duplicates in our FCs

```{r}
duplicates_of_log2FCs = duplicated(log2FC)
length(duplicates_of_log2FCs[duplicates_of_log2FCs == TRUE]) # we have 10 out of 7801 duplicates
```

load the fGSEA package

```{r}
library(fgsea)
```


```{r}
rank_dat = dat_for_ranking$log2FC
names(rank_dat) = dat_for_ranking$gene_ens_ID
head(rank_dat)
rank_dat = data.frame(rank_dat)
dim(rank_dat)
colnames(rank_dat) = c("log2FC")
```
rank the fold change while randomly ranking the ties

```{r}
rank_dat$untied = rank(rank_dat$log2FC, ties.method = "random")
rank_dat = rank_dat %>% arrange(desc(untied))
ranks = rank_dat$untied
names(ranks) = dat_for_ranking$gene_ens_ID
head(ranks)

```

```{r}
total_pathways_ensID <- readRDS("~/GitHub/2022-topic-02-team-04/data/total_pathways_ensID.rds")
fgsea_dat = fgsea(total_pathways_ensID,
                  stats = ranks,
                  minSize =1, 
                  maxSize = 1000, scoreType = "pos")

```

```{r}
fgsea_dat
```

```{r}
plotEnrichment(total_pathways_ensID[["spermatogenesis"]], ranks)
```

```{r}
plotEnrichment(total_pathways_ensID[["apoptosis"]], ranks)
```

```{r}
library("BiocManager")
library("fgsea")
library("data.table")
library("ggplot2")
library("dplyr")
topUp <- fgsea_dat %>% 
    filter(ES > 0) %>% 
    top_n(10, wt=-padj)
topDown <- fgsea_dat %>% 
    filter(ES < 0) %>% 
    top_n(10, wt=-padj)
topPathways <- bind_rows(topUp, topDown) %>% 
    arrange(-ES)

plotGseaTable(total_pathways_ensID[topPathways$pathway], 
              ranks, 
              fgsea_dat, 
              gseaParam = 0.5)
```

What kind of significant pathways do we have?

```{r}
sig_pathways_Gsea = fgsea_dat[which(fgsea_dat$ES > 0 & fgsea_dat$pval < 0.05),] # only two pathways "Meta_Chol" and "cholesterol homeostasis"
```