---
title: "Linear regression - predicting whether we have cancer or normal cell samples"
author: "Lottida Phondeth, Carla Welz"
date: "`r format(Sys.time(),'%d %B %Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load data

```{r}
setwd('..')
tcga_anno = readRDS("~/GitHub/2022-topic-02-team-04/data/tcga_tumor_annotation.RDS") 
PRAD_TvsN = readRDS("~/GitHub/2022-topic-02-team-04/data/PRAD_TvsN.rds")
pw_activity_full_TCGA_df = readRDS("~/GitHub/2022-topic-02-team-04/data/pw_activity_full_TCGA_df.rds")
tcga_pancancer = readRDS("~/GitHub/2022-topic-02-team-04/data/tcga_pancancer.rds")
tcga_anno = readRDS("~/GitHub/2022-topic-02-team-04/data/tcga_tumor_annotation.RDS")
tcga_tumor_norm = readRDS("~/GitHub/2022-topic-02-team-04/data/PRAD_tumor_normal_pw_activity.rds")

```

# Objectives

In this task it is our goal to generate a multiple regression model that can determine whether the obtained cell samples are cancerous or not.
We might have to determine the pathway that can be used to predict the sample state.

First we generated our test and training sample by dividing our pan cancer dataset and out tumor vs normal dataset. 

```{r}
PRAD_pan = as.data.frame(pw_activity_full_TCGA_df[which(tcga_anno$cancer_type_abbreviation == "PRAD" )])

PRAD_tumor = as.data.frame(tcga_tumor_norm[,1:52])

PRAD_normal = as.data.frame(tcga_tumor_norm[,53:104])

PRAD_pan = subset(PRAD_pan,rownames(PRAD_pan) %in% rownames(PRAD_tumor) )

test_tumor = PRAD_tumor[,1:round(ncol(PRAD_tumor)*0.4,0)]

test_normal = PRAD_normal[,1:round(ncol(PRAD_normal)*0.4,0)]

test_pan = PRAD_pan[,1:round(ncol(PRAD_pan)*0.4,0)]

train_tumor = PRAD_tumor[,(round(ncol(PRAD_tumor)*0.4,0)+1):ncol(PRAD_tumor)]

train_normal = PRAD_normal[,(round(ncol(PRAD_normal)*0.4,0)+1):ncol(PRAD_normal)]

train_pan =  PRAD_pan[,(round(ncol(PRAD_pan)*0.4,0)+1):ncol(PRAD_pan)]

dat.train = cbind(train_tumor, train_normal, train_pan)

dat.test = cbind(test_tumor, test_normal, test_pan)
```

The variance of the pathways was computed, sorted accordingly and afterwards 100 pathways with the highest variance were selected.

```{r}
library(dplyr)
library(pheatmap)
variance <- as.data.frame(apply(dat.train, 1, var))
colnames(variance) = "variance"

var_sorted <- arrange(variance, desc(variance))

## take 9 pws mit highes variance
topVar <- rownames(var_sorted)[1:100]

## take these pws
topVar_pw <- as.data.frame(dat.train[topVar,])

##
topVar_cor <- cor(t(topVar_pw)); diag(topVar_cor) = NA
pheatmap(topVar_cor,
         show_colnames = F,
         show_rownames = F,
         cluster_rows = FALSE,
         cluster_cols = FALSE)

```

Here we select the pathways with the highest variance (TopVar100)
```{r}
b = data.frame()
lr_pw = c( )
lr_pw = c(rownames(topVar_pw[1,]))
 for(p in 2:100){
   correlation = FALSE
   for(f in 1:length(lr_pw)){
     if(p!=f){
      if(topVar_cor[f,p]>0){
      correlation = TRUE
      }
    }
   }
   b[p,1] = correlation
   if(correlation == FALSE){
   lr_pw = c(lr_pw,rownames(topVar_pw)[p])
   }
 }
```

Generate the annotations of the patients (cancer vs no cancer)

```{r}
#annotations for each patient group
tumor_anno <- as.data.frame(rep("tumor", 52))
colnames(tumor_anno) = "sample"
rownames(tumor_anno) = colnames(tcga_tumor_norm[,1:52])
tumor_anno = t(tumor_anno)

normal_anno <- as.data.frame(rep("normal", 52))
colnames(normal_anno) = "sample"
rownames(normal_anno) = colnames(tcga_tumor_norm[,53:104])
normal_anno = t(normal_anno)

pan_anno <- as.data.frame(rep("tumor", ncol(PRAD_pan)))
colnames(pan_anno) = "sample"
rownames(pan_anno) = colnames(PRAD_pan)
pan_anno = t(pan_anno)
#divide annotations for test and training set
test_tumor_anno = as.data.frame(tumor_anno[,1:round(ncol(PRAD_tumor)*0.4,0)])
test_tumor_anno = t(test_tumor_anno)
#colnames(test_tumor_anno) = colnames(tumor_anno[1:round(ncol(PRAD_tumor)*0.4,0)])
test_normal_anno = as.data.frame(normal_anno[,1:round(ncol(PRAD_normal)*0.4,0)])
test_normal_anno = t(test_normal_anno)

test_pan_anno = as.data.frame(pan_anno[,1:round(ncol(PRAD_pan)*0.4,0)])
test_pan_anno = t(test_pan_anno)

train_tumor_anno = as.data.frame(tumor_anno[,(round(ncol(PRAD_tumor)*0.4,0)+1):ncol(PRAD_tumor)])
train_tumor_anno = t(train_tumor_anno)

train_normal_anno = as.data.frame(normal_anno[,(round(ncol(PRAD_normal)*0.4,0)+1):ncol(PRAD_normal)])
train_normal_anno = t(train_normal_anno)

train_pan_anno =  as.data.frame(pan_anno[,(round(ncol(PRAD_pan)*0.4,0)+1):ncol(PRAD_pan)])
train_pan_anno = t(train_pan_anno)

dat.train_anno = cbind(train_tumor_anno, train_normal_anno, train_pan_anno)

dat.test_anno = cbind(test_tumor_anno, test_normal_anno, test_pan_anno)
```

```{r}
topVar_pw_anno = as.data.frame(t(topVar_pw))
topVar_pw_anno = cbind(topVar_pw_anno, t(dat.train_anno))
colnames(topVar_pw_anno[101])="anno"

```

```{r}
lr = lm( ~ HINATA_NFKB_IMMU_INF, data=topVar_pw_anno)
summary(lr)

hist(lr$residuals)
qqnorm(lr$residuals); qqline(lr$residuals)
cor(topVar_pw$HALLMARK_G2M_CHECKPOINT, lr$residuals)
plot(topVar_pw$HALLMARK_G2M_CHECKPOINT, lr$residuals)
```

